---
title: "exercise-11"
author: "Erika Carlson"
date: 2023-03-25
format: html
editor_options: 
  chunk_output_type: console
---

# Exercise 11 {.unnumbered}

# Practice ANOVA {.unnumbered}

#### Preliminaries {.unnumbered}

> **Data source**:
>
> Tobias JA, et al. (2022). AVONET: Morphological, ecological and geographical data for all birds. *Ecology Letters* 25: 581–597.

* Load in Morphological measurements on birds, AVONET data set
  + This is a recently published dataset that compiles morphological measurements and information on various ecological variables and geographic range data for more than 11,000 species of birds.

* Winnow the dataset to include only the following variables: **Species1**, **Family1**, **Order1**, **Beak.Length_Culmen**, **Beak.Width**, **Beak.Depth**, **Tarsus.Length**, **Wing.Length**, **Tail.Length**, **Mass**, **Habitat**, **Migration**, **Trophic.Level**, **Trophic.Niche**, **Min.Latitude**, **Max.Latitude**, **Centroid.Latitude**, **Primary.Lifestyle**, and **Range.Size**

```{r}
#| warning: false
library(tidyverse)
library(broom)

d <- read_csv("https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv", col_names = TRUE)

# winnow data set
d <- d %>% select(Species1, Family1, Order1, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Range.Size, Beak.Length_Culmen, Primary.Lifestyle)
```

* Run exploratory data analysis with this dataset and determine which variables are categorical and which are numeric.

```{r}
glimpse(d)
```

  + *Species, Family, Order, Habitat are categorical and character*
  + *Migration is also categorical but numeric*
  + *All others are numeric*


## Challenge 1 {.unnumbered}
### One-Factor ANOVA and Inference {.unnumbered}

#### Step 1 {.unnumbered}

* Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, convert the variable **Migration** (which is scored as a number: "1", "2", or "3") from class numeric to either being classified as a factor or as a character (string) variable.

```{r}
# plot Mass as a factor of Trophic Level
p1 <- ggplot(data = d %>% drop_na(Trophic.Level),
             aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot() +
  geom_jitter(alpha = 0.05) +
  xlab("Trophic level")

# plot Mass as a factor of Migration (include argument as.factor())
p2 <- ggplot(data = d %>% drop_na(Migration),
             aes(x = as.factor(Migration), y = log(Mass))) +
  geom_boxplot() +
  geom_jitter(alpha = 0.05) +
  xlab("Migration")

library(ggpubr)
ggarrange(p1, p2,
          labels = NULL,
          ncol = 2, nrow = 1)
```


#### Step 2 {.unnumbered}

* Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

```{r}
m1 <- lm(log(Mass) ~ Trophic.Level, data = d)

# need to coerce Migration as a factor before modeling
d <- d %>% mutate(Migration = as.factor(Migration))
m2 <- lm(log(Mass) ~ Migration, data = d)
```

* Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration** category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

```{r}
# check beta coefficients
summary(m1) # low p value for F statistic, very low adjusted R squared
# Estimate tells us that for a change in trophic level from Carnivore (base) to Omnivore, there is a 25% increase in mean body mass
# The coefficient for the intercept, i.e. beta_0 reflects the estimate of the mean value for Carnivores

summary(m2)
# migration
# sedentary = 1
# partially migratory = 2
# migratory = 3
```

  + *Yes, log(Mass) is associated with both Trophic.Level and Migration, as both models have large F-statistics (78.42 and 144.5) and very low associated p-values (< 2.2e-16 for both)*
* Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level?
  + *Partially migratory and migratory are different than sedentary, level 1.*
* Relevel and assess differences among the remaining pair of **Migration** categories.
 
```{r}
m2 <- lm(log(Mass) ~ Migration, data = d %>% mutate(Migration = relevel(Migration, ref = "3")))
summary(m2)
```

  + *Partially migratory and Migratory are also different*
* The coefficient for the intercept, beta_0, reflects the estimate of the mean log(Mass) for the first of our level variables
* The estimate of beta_1 is the estimated difference in mean log(Mass) associated with being that level relative to the first (reference) level


#### Step 3 {.unnumbered}

* Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ "significantly" from one another (see [**Module 20**](#module-20)).
  + After finding a significant omnibus F statistic in an ANOVA, we can test, post-hoc, what group means are different from on another using pairwise t-tests with an appropriate p value correction or a Tukey Honest Significant Differences test

```{r}
m1 <- lm(log(Mass) ~ Trophic.Level, data = d)

summary(m1)

# alternative way to run simple linear regression with `aov()`
m1aov <- aov(log(Mass) ~ Trophic.Level, data = d) 

summary(m1aov)

# Pairwise posthoc t-test
(pairwise.t.test(log(d$Mass), d$Trophic.Level, p.adj = "bonferroni")) 

# Tukey test runs on aov object and is a comparable test, run slightly different
(posthoc <- TukeyHSD(m1aov, which = "Trophic.Level", conf.level = 0.95))
```


#### Step 4 {.unnumbered}

* Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic using the {infer} workflow and setting `calculate(stat="F")`.

```{r}
library(infer)

# calculate the original F statistic
original.F <- aov(log(Mass) ~ Trophic.Level, data = d) %>% 
  tidy() %>% # pull table of coefficients from broom package
  filter(term == "Trophic.Level") # assign results of specific variable
original.F # results for F stat and p value for omnibus F test

# generate a permutation distribution, breaking the association between predictor and response variable
# generate a separate variable for log of body mass
d <- d %>% mutate(logMass = log(Mass))

permuted.F <- d %>% 
  specify(logMass ~ Trophic.Level) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "F")
permuted.F

hist(permuted.F$stat) # zero bounded, with decay (looks like an F distribution!)

visualize(permuted.F) +
  shade_p_value(obs_stat = original.F$statistic, 
                direction = "greater") # very far away from distribution because p value associated is very very tiny
```

Statistics is always comparing an observed test statistic to a distribution (whether a standardized distribution OR permuted distribution of test statistics from your data set)

* The number of times you generate a test statistic under null that the actual value of test statistic exceeds that permuted value
* 1000 permutations is usually good, maybe 10000


## Challenge 2 {.unnumbered}
### Data Wrangling, One- and Two-Factor ANOVA {.unnumbered}

#### Step 1 {.unnumbered}
* Create the following two new variables and add them to AVONET dataset:
  + **Relative beak length** = the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).
  + **Relative tarsus length** = the *residual* of log(**Tarsus.Length**) on log(**Mass**).

```{r}
# failed strategy for pulling residuals
# 
# m2 <- lm(logMass ~ log(Beak.Length_Culmen), data = d)
# 
# summary(m2)
# 
# d <- d %>% mutate(relative_beak_length = m2$residuals)
# 
# p <- ggplot(data = d %>% drop_na(Primary.Lifestyle), aes(x = Primary.Lifestyle, y = relative_beak_length)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))
# p

d$relBL <- resid(lm(formula = log(Beak.Length_Culmen) ~ logMass, data = d))

d$relTL <- resid(lm(formula = log(Tarsus.Length) ~ logMass, data = d))
```


#### Step 2 {.unnumbered}

* Make a boxplot of relative tarsus length variable in relation to **Primary.Lifestyle** and of relative beak length variable in relation to **Trophic.Niche**

```{r}
p1 <- ggplot(data = d %>% drop_na(Primary.Lifestyle), 
             aes(x = Primary.Lifestyle, y = relTL)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Primary lifestyle") + ylab("Relative tarsus length")

p2 <- ggplot(data = d %>% drop_na(Trophic.Niche), 
             aes(x = Trophic.Niche, y = relBL)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Trophic niche") + ylab("Relative beak length")

ggarrange(p1, p2,
          labels = NULL,
          ncol = 2, nrow = 1)
```


#### Step 3 {.unnumbered}

* Run an ANOVA analyses to look at the association between geographic range size and the variable **Migration**
  + Drop those observations for which **Migration** is not scored
  + Look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed (log)
  + Based on the global model, is range size associated with form of migration? *Yes*
  + How much of the variance in your measure of range size is associated with **Migration** behavior style? *Adjusted r-squared = 0.08674, so 8.7% of the variance*

- Given the regression coefficients returned in output of the model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ "significantly" from one another (see [**Module 20**](#module-20)).

Run an ANOVA on geographic range as a function of migratory status

```{r}
m1aov <- aov(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration)) 
summary(m1aov)
TukeyHSD(m1aov)

ggplot(d %>% drop_na(Migration), 
       aes(x = as.factor(Migration), y = log(Range.Size))) +
  geom_boxplot() +
  geom_jitter(alpha = 0.05) +
  xlab("Migration") + ylab("log(Range size)")

# get the R-squared of the anova by looking at the summary output of the corresponding linear model
m1 <- lm(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration))
anova(m1)
summary(m1)
```


#### Step 4 {.unnumbered}

- Winnow the data to just consider birds from the Infraorder "Passeriformes" (song birds).

```{r}
d_pass <- d %>% filter(Order1 == "Passeriformes")
```

* Run separate one-factor ANOVA analyses to look at the association between [1] relative beak length and **Primary.Lifestyle** and between [2] relative beak length and **Trophic.Level**. 
  + Make boxplots of response variable by each predictor and by the combination of predictors.
  + Run linear models for each predictor separately and interpret the model output.
  + *Primary lifestyle and trophic level significantly effect relative beak length when modeled separately, based on large F values and very small p-value associated with each global model*
  
```{r}
m0aov <- aov(relBL ~ 1, data = d_pass)

m1aov <- aov(relBL ~ Primary.Lifestyle, data = d_pass)
summary(m1aov)

m2aov <- aov(relBL ~ Trophic.Level, data = d_pass)
summary(m2aov)

p1 <- ggplot(data = d_pass %>% drop_na(Primary.Lifestyle), 
             aes(x = Primary.Lifestyle, y = relBL)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.025) +
  xlab("Primary lifestyle") + ylab("Relative beak length")

p2 <- ggplot(data = d_pass %>% drop_na(Trophic.Level), 
             aes(x = Trophic.Level, y = relBL)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.025) +
  xlab("Trophic level") + ylab("Relative beak length")

# combination
p3 <- ggplot(data = d_pass %>% drop_na(Primary.Lifestyle), 
             aes(x = Primary.Lifestyle, y = relBL)) + 
  geom_boxplot() + 
  facet_wrap(~ Trophic.Level) +
  geom_jitter(alpha = 0.025) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Primary lifestyle") + ylab("Relative beak length")

library(ggpubr)
ggarrange(p1, p2, p3,
          labels = NULL,
          ncol = 2, nrow = 2)
```


#### Step 5 {.unnumbered}

* Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms. 
* Based on the model output, what would you conclude about how relative beak length is related to these two variables? *Both Trophic.Level and Primary.Lifestyle are predictive of relative beak length, and this combination accounts for 10.9% of the variance in relative beak length*

```{r}
m1aov2 <- aov(relBL ~ Trophic.Level + Primary.Lifestyle, 
              data = d_pass)
summary(m1aov2)

# get the R-squared of the anova by looking at the summary output of the corresponding linear model
m1 <- lm(relBL ~ Trophic.Level + Primary.Lifestyle, 
              data = d_pass)
summary(m1)
```


#### Step 6 {.unnumbered}

* Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term using the colon operator (`:`)
  + e.g., relative beak length ~ **Primary.Lifestyle** + **Trophic.Level** + **Primary.Lifestyle:Trophic.Level**. 
  + Based on the model output, what would you now conclude about how relative beak length is related to these two variables? *Both Trophic.Level and Primary.Lifestyle are predictive of relative beak length, as is the interaction between these two predictors, all accounting for 13.5% of the variance in beak length*

```{r}
m2aov2 <- aov(relBL ~ Trophic.Level + Primary.Lifestyle + Trophic.Level:Primary.Lifestyle, 
              data = d_pass)
summary(m2aov2)

m2 <- lm(relBL ~ Trophic.Level + Primary.Lifestyle + Trophic.Level:Primary.Lifestyle, 
              data = d_pass)
summary(m2)
```

```{r}
# can do more variables using *
m1aov3 <- aov(relBL ~ Trophic.Level * Primary.Lifestyle * as.factor(Migration), 
              data = d_pass)
```

#### Step 7 {.unnumbered}

- Use the `interaction.plot()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level** (see [**Module 20**](#module-20)).

```{r}
interaction.plot(
  x.factor = d_pass$Trophic.Level,
  xlab = "Trophic level",
  trace.factor = d_pass$Primary.Lifestyle,
  trace.label = "Primary lifestyle",
  response = d_pass$relBL,
  ylab = "Mean relative beak length"
)
```

#### Additional Steps? {.unnumbered}
(in progress)

In the exercise above, we really did not do any checking with this dataset to see if the data meet the primary assumptions for standard linear regression and ANOVA, which are that variables/residuals within each grouping level are roughly normally distributed and have roughly equal variances. Sample sizes within each grouping level should also be roughly equal. As noted in [**Module 20**](#module-20), a general rule of thumb for "equal" variances is to compare the largest and smallest within-grouping level standard deviations and, if this value is less than 2, then it is often reasonable to presume the assumption may not be violated.

Use this approach to see whether variances in across groups in your various models (e.g., for relative beak length ~ trophic level) are roughly equal. Additionally, do a visual check of whether observations and model residuals within groups look to be normally distributed.

* Check residual analysis by eye

Looking for a nice scatter of points around a horizontal line (should be no relationship)
```{r}
#| eval: false
library(car)
install.packages("jtools")
library(jtools)

plot(m1$model$Trophic.Level, residuals(m1))
plot(m1$model$Primary.Lifestyle, residuals(m1))
plot(fitted(m1), residuals(m1)) # plotting residuals given combinations of Trophic.Level and Primary.Lifestyle
plot(m1) # opens interactive plot window
```
